### 2017-05-29 src/5_1_Proposal.tex
adding last changes from Springer journal

\chapter{Adaptive parallel algorithm for Vector error correction model}
VECM (presented in section \ref{sec:vecm}) has been
broadly used with low frequency time series. We aimed to adapt VECM to be used
in finance with high frequency stream data.

We observed that the number of cointegration relationships varies with the
length of historical data used. Moreover, parameters that increased these
relationships in time led to better forecasting performance. Our proposal,
called an Adaptive VECM (AVECM) is to make a parameters grid search that
maximises the number of cointegration relationships in the near past. To ensure
the search can be executed fast enough, we used a distributed environment.

@@

\section{Introduction}
\label{sec:introduction}
In finance, it is common to find variables with long-run equilibrium
relationships. This is termed cointegration and reflects the idea that some sets
of variables cannot wander too far from each other. Cointegration means that one
or more linear combinations of these variables are stationary even though
individually they are not. Some models, such as the Vector Error Correction
(VECM), see \cite{engle87}, take advantage of this property and describe the
joint behaviour of several cointegrated variables. VECM introduces this long-run
relationship among a set of cointegrated time series as an error correction
term. These time series must be integrated of order 1, denoted I(1), i.e. they
become stationary at their first differences. In finance, I(1) time series are
very common and to introduce cointegration restrictions in models often improves
forecasting, see \cite{duy1998}. Therefore, VECM has been widely adopted in
financial applications, among others: \cite{mukherjee1995}, \cite{seong2013},
\cite{maysami2000} and \cite{arestis2001}. VECM has also been used in pair
trading, see \cite{herlemont2003}, or models with more than two variables, see
for example \cite{mukherjee1995} and \cite{engle2004}).

VECM parameters are obtained using the ordinary least squares (OLS) method,
developed by \cite{golub1980}. Since OLS involves many calculations, the
parameter estimation is computationally expensive when the number of lagged
values and data increases. Moreover, obtaining cointegration vectors is also an
expensive routine. This is a main limitation to use VECM with stream data with
high frequency. \cite{chen2003} addresses the advantage of distributed
processing over conventional rolling window processing.

Our aims were to study if a parallel version of VECM can be used with high
frequency stream data.  Our approach was to determine adaptively the number of
observations and lags of VECM which maximise cointegration relations in the
past.  This search is computationally expensive because the Johansen method, see
\cite{johansen1995}, is required to find cointegration vectors, which is of
order $O(n^3)$. Therefore, our proposal, called Adaptively Vector Error
Correction (AVECM), is to parallelise this search in order to get new parameters
before new data arrives.  Model effectiveness is focused on out-of-sample
forecast rather than in-sample fitting. This criterion allows AVECM prediction
capability to be expressed rather than just explaining data history.  The
forecast capability of our method was measured using MSE and the Theil's
$U$-statistic, see \cite{theil1966}, widely used in economic forecast. Tests
were run using four currency rates: Euro (EUR) to United States Dollar (USD)
(EURUSD), British Pound (GBP) to USD (GBPUSD), USD to Swiss Franc (CHF) (USDCHF)
and USD to Japanese Yen (JPY) (USDJPY) with a 10-second frequency.

@@

Cointegration vectors can be found applying the Johansen method which uses a
sample of the last historical data. However, VECM assumes cointegration vectors
do not change in time.  In fact, \cite{gregoryETal1996} addresses that the
long-run relationships between the time series  might change due to several
economic factors that can lead to structural breaks in the cointegration
relationship.  In order to show that the number of cointegration vectors depends
on the amount $L$ of historical data and the number of lags $p$ in the VECM, we
used a grid search.  We defined a grid of possible values for $L$ and $p$. $L$
goes throughout $[2,14]$ hours (1 $hour = 360$ data points) with a step size of
4 hours and $p$ throughout $[1,5]$ with step size of 1. The first experiment
consisted in determining the number of cointegration vectors for all
combinations of $L$ and $p$.  We used four forex rates: EURUSD, GBPUSD, USDCHF
and USDJPY with 10-second frequency. Data started at 13:00 GMT of the 13th of
August 2014, when the New York and London financial markets opened.

Figure \ref{fig:hists} shows the distribution of the number of cointegration
vectors given by the Johansen method for different values of $L=[2,6,10,14]$
hours  and $p=1$ . This procedure was carried out by a sliding window of
historical data moving 1000 times.

\begin{figure}[!h]
  %\vspace{-0.8cm}
  \centering
  \includegraphics[width=0.9\textwidth]{img/Fig1}
  \caption{Distribution of the number of cointegration vectors using $p=1$ lags.
  Four possible values for windows size $L$ are shown (2, 6, 10 and 14) hours (1
  hour = 360 data points).}
  \label{fig:hists}
\end{figure}

@@

The goal of our next experiment was to find a relation between this ratio $PC$
and the performance measure MSE (see equation \ref{eq:MSE}). $L$ was defined
between $[700,1500]$ data points with step size 100 and $p$ took values between
$[1,5]$ with step size 1.

\begin{figure}[ht!]
  \centering
  \includegraphics[width=0.9\textwidth]{img/Fig2}
  \caption{MSE versus the percentage of cointegration considering 1000
  iterations. Best windows size $L$ found was 1060. Below MSE versus $L$ shows a
  rapidly decreasing behaviour, founding minimum at PC = 80\% }
  \label{fig:cointvsmse}
\end{figure}

Figure~\ref{fig:cointvsmse} shows the relation between MSE and $PC$ and $L$.  We
found that better cointegration percentage leads to better performance accuracy
in terms of reducing MSE.

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.9\textwidth]{img/Fig3}
  \caption{Computing time of sequential and parallel algorithm is shown in the
  upper figure. Speed-up is shown below.}
  \label{fig:extimes}
\end{figure}

>>>>

\chapter{Fast and adaptive cointegration based model
for forecasting high frequency financial time series}

Cointegration is a long-run property of some non-stationary time series where a
linear combination of those time series is stationary.  This behaviour has been
studied in finance because cointegration restrictions often improve forecasting.
The Vector Error Correction Model (VECM) is a well-known econometric technique
that characterises short-run variations of a set of cointegrated time series
incorporating long-run relationships as an error correction term. VECM has been
broadly used with low frequency time series. We aimed to adapt VECM to be used
in finance with high frequency stream data.

Cointegration relations change in time and therefore VECM parameters must be
updated when new data is available. We studied how forecasting performance is
affected when VECM parameters and the length of historical data used change in
time. We observed that the number of cointegration relationships varies with the
length of historical data used. Moreover, parameters that increased these
relationships in time led to better forecasting performance. Our proposal,
called an Adaptive VECM (AVECM) is to make a parameters grid search that
maximises the number of cointegration relationships in the near past. To ensure
the search can be executed fast enough, we used a distributed environment.

@@

\section{Introduction}
\label{sec:introduction}
In finance, it is common to find variables with long-run equilibrium
relationships. This is termed cointegration and reflects the idea that some sets
of variables cannot wander too far from each other. Cointegration means that one
or more linear combinations of these variables are stationary even though
individually they are not. Some models, such as the Vector Error Correction
(VECM), see \cite{engle87}, take advantage of this property and describe the
joint behaviour of several cointegrated variables. VECM introduces this long-run
relationship among a set of cointegrated time series as an error correction
term. These time series must be integrated of order 1, denoted I(1), i.e. they
become stationary at their first differences. In finance, I(1) time series are
very common and to introduce cointegration restrictions in models often improves
forecasting, see \cite{duy1998}. Therefore, VECM has been widely adopted in
financial applications, among others: \cite{mukherjee1995}, \cite{seong2013},
\cite{maysami2000} and \cite{arestis2001}. VECM has also been used in pair
trading, see \cite{herlemont2003}, or models with more than two variables, see
for example \cite{mukherjee1995} and \cite{engle2004}.

Cointegration relationships can be found in low and high frequency data of two or
more assets. While cointegration in low frequency data is motivated by a
long-run equilibrium relationship between economic forces, cointegration in high
frequency data has its foundation in statistical arbitrage theory which is very helpful
to detect mean-reverting trades. Information about cointegrated assets in high
frequency data could be used as an input for high frequency trading strategies,
using it as a signal to capture small profits in short term trades, see
\cite{miao2014}. \cite{zhou2001} addressed the benefits of using higher
frequency data to analize cointegration. \cite{rittler2012} also found that
cointegration may differ with different data frequencies and could appear
whith increased data frequency.

The use of VECM with high frequency data is mainly limited by computationally
expensive routines. Firstly, VECM parameters are obtained using the ordinary
least squares (OLS) method, developed by \cite{golub1980}. Since OLS involves
many calculations, the parameter estimation is computationally expensive when
the number of lagged values and data increases. Secondly, obtaining
cointegration vectors is also an expensive routine because the Johansen method 
is required, which is of order $O(n^3)$ (\cite{johansen1995}).
\cite{chen2003} addressed the advantage of distributed processing over
conventional rolling window processing. Therefore, our aim was to study if a
parallel version of VECM can be used with high frequency stream data.

Our approach was to determine, adaptively, the number of observations and lags of
VECM which maximise cointegration relations in the past in a distributed
environment. We called our proposal Adaptive Vector Error Correction (AVECM). 
AVECM parallelises this search of parameters in order to update them
before new data arrives. Model effectiveness is focused on out-of-sample
forecast rather than in-sample fitting. This criterion allows AVECM prediction
capability to be expressed rather than just explaining data history.  The
forecast capability of our method was measured using MSE and the Theil's
$U$-statistic, see \cite{theil1966}, widely used in economic forecast. Tests
were run using four currency rates: Euro (EUR) to United States Dollar (USD)
(EURUSD), British Pound (GBP) to USD (GBPUSD), USD to Swiss Franc (CHF) (USDCHF)
and USD to Japanese Yen (JPY) (USDJPY) with a 10-second frequency.

This paper is organised as follows: section \ref{sec:background} presents the
VAR and VECM, the AVECM algorithm is presented in section \ref{sec:methodology}.
In section \ref{sec:results} we describe the tests carried on to assess the
accuracy and the execution time of AVECM.  This section also includes a
description of the test data.  Section \ref{sec:conclusions} contains the
conclusions and a discussion of future research.

\section{Background}
\label{sec:background}

\subsection{Integration and Cointegration}\label{sec:coint}\  
Following \cite{johansen1995} we shall say that a stochastic process
$Y_t$ which satisfies $Y_t-E(Y_t) = \sum_{i=0}^\infty C_i\,\varepsilon_{t-i}$ is
called $I(0)$, and then we shall write $Y_t\sim I(0)$, whenever
$\sum_{i=0}^\infty C_i \neq 0$ and $\sum_{i=0}^\infty C_i\,z^i$ converges for
$z\in\mathbb{C}$ with $|z|<1$.  It is understood that the condition
$\varepsilon_t\sim iid(0,\sigma^2)$ holds.

A (vector) time series $\mathbf{y}_t$ is said to be {\em integrated of order\/}
$d$, and then we shall write $\mathbf{y}_t\sim I(d)$, whenever after $d$ times
(discrete) differentiation an stationary process is
obtained, see \cite{banerjee1993}; more precisely, whenever
$(1-L)^d\,\mathbf{y}_t\sim\text{I(0)}$, where $L$ is the usual lag operator:
$(1-L)\,\mathbf{y}_t = \Delta\mathbf{y}_t = \mathbf{y}_t-\mathbf{y}_{t-1}$ for
all $t$.

Note that this definition includes the scalar case as time series of vectors of
dimension 1; in this scalar case we will write the time series in nonbold
format.

Let $\mathbf{y}_t^\nu$, $\nu=1,\dots,l$, be a set of $l$ vector time series of
order $I(1)$.  They are said to be {\em cointegrated\/} if a vector
$\beta=[\beta(1),\dots,\beta(l)]^\top \in \mathbb{R}^p$ exists, such that the
time series,
\begin{equation}
\mathbf{Z}_t:= 
\sum_{\nu=1}^l \beta(\nu)\,\mathbf{y}_t^\nu\,\sim\,\text{I(0)}\,.
\end{equation}
In other words, a set of $I(1)$ time series is said to be cointegrated if a
linear combination of them exists, which is I(0).

\subsection{Vector Autorregresive Models}\label{sec:varvec}

Vector error correction model (VECM) describe the joint behaviour of a set of
variables and can be derived from the simple Vector Autoregressive model (VAR)
presented by \cite{sims1980}.  The VAR($p$) model is a framework describing the
behaviour of a set of $l$ endogenous and stationary variables as a linear
combination of their last $p$ values, where $l,p\in\mathbb{N}$.  In our case,
each one of these $l$ variables is a scalar time series $y_{\lambda,t}$,
$\lambda=1,\dots,l$, and we represent them all together at time $t$ by the
vector time series:
\begin{equation}
\label{eq:variables}
\mathbf{y}_t = 
\begin{bmatrix} y_{1,t} & y_{2,t} & \dots & y_{l,t} \end{bmatrix}^\top.
\end{equation}
\noindent
Notice that the vectors $\mathbf{y}_t$ are assumed to be $l$-dimensional.

The VAR($p$) model describes the behaviour of a dependent variable in terms of
its own lagged values and the lags of the others variables in the system. The
model with $p$ lags is formulated as the system:
\begin{align}
\label{eq:var}
\mathbf{y}_t 
= \boldsymbol{\Phi}_1 \mathbf{y}_{t-1} +
  \boldsymbol{\Phi}_2 \mathbf{y}_{t-2} + \dots +
  \boldsymbol{\Phi}_p\mathbf{y}_{t-p} +
  \mathbf{c} + \boldsymbol{\epsilon}_t \nonumber \\
t=p+1,\dots,N,
\end{align}
\noindent where 
$\boldsymbol{\Phi}_1, \boldsymbol{\Phi}_2,\dots,\boldsymbol{\Phi}_p$
are $l\times l$-matrices of real coefficients,
$\boldsymbol{\epsilon}_{p+1},
 \boldsymbol{\epsilon}_{p+2}, \dots, \boldsymbol{\epsilon}_N$ 
are error terms, $\mathbf{c}$ is a constant vector and $N$ is the total
number of samples.

Notice that, regarding our notation of section (\ref{sec:coint}),
we have here 
$\mathbf{y}_t^0 = \mathbf{y}_t$,
$\mathbf{y}_t^\nu = \mathbf{y}_{t-\nu}$ and
the $\lambda$-th component of the vector time series $\mathbf{y}_t^\nu$
is the scalar time series $y_{\lambda,t}^\nu$, where $\nu=1,\dots,p$ and
$\lambda=1,\dots,l$.

However, the VAR model cannot be used with non-stationary variables. VECM ,
developed by \cite{engle87}, is also a linear model for I(1) variables that are
also cointegrated, see \cite{banerjee1993}. If cointegration exists, variable
differences are stationary and they introduce an error correction term which
adjusts coefficients to bring the variables back to equilibrium.

It is obtained re-writing equation (\ref{eq:var}) in terms of the new
variable $\Delta\mathbf{y}_t=\mathbf{y}_t-\mathbf{y}_{t-1}$.
The VECM model, expressed in terms those differences, takes the form:
\begin{equation}\label{eq:vec}
\Delta \mathbf{y}_t 
= \boldsymbol{\Omega}\,\mathbf{y}_{t-1}
  + \sum_{i=1}^{p-1} \boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
\end{equation}
\noindent
where the coefficients matrices $\boldsymbol{\Phi}_i^*$ and 
$\boldsymbol{\Omega}$, expressed in terms of the matrices
$\boldsymbol{\Phi}_i$ of (\ref{eq:var}), are:
\begin{align*}
\boldsymbol{\Phi}_i^* 
&:= -\sum_{j=i+1}^{p}\boldsymbol{\Phi}_j\,, \\
\boldsymbol{\Omega}
&:= -\left( \mathbb{I} - \boldsymbol{\Phi}_1 - \dots 
    - \boldsymbol{\phi}_p \right)\,. 
\end{align*}
The following well known properties of the matrix $\boldsymbol{\Omega}$, see
\cite{johansen1995}, will be useful in the sequel:
\begin{itemize}
\item
If $\boldsymbol{\Omega} = \mathbf{0}$, there is no cointegration.
\item 
If $rank(\boldsymbol{\Omega})=l$, i.e., if $\boldsymbol{\Omega}$ has
full rank, then the time series are not I(1) but stationary.
\item
If $rank(\boldsymbol{\Omega})=r$, $0<r<l$, then there is cointegration
and the matrix $\boldsymbol{\Omega}$ can be expressed as
$\boldsymbol{\Omega}=\boldsymbol{\alpha\beta}^\top$, where $\boldsymbol{\alpha}$
and $\boldsymbol{\beta}$ are
$l\times r$ matrices and
$\text{rank}(\boldsymbol{\alpha})=\text{rank}(\boldsymbol{\beta})=r$.
\item
The columns of $\boldsymbol{\beta}$ contains the cointegration vectors and the rows of
$\boldsymbol{\alpha}$ correspond with the adjusted vectors. 
$\boldsymbol{\beta}$ is obtained by Johansen procedure,see \cite{johansen1988},
whereas $\boldsymbol{\alpha}$ has to be determined as a variable in the VECM.
\end{itemize}
It is worth noticing that the factorization of the matrix
$\boldsymbol\Omega$ is not unique, since for any $r \times r$
non-singular matrix $\mathbf{H}$, $\boldsymbol{\alpha}^*:=\boldsymbol{\alpha}\mathbf{H}$,
and $\boldsymbol{\beta}^*=\boldsymbol{\beta}(\mathbf{H}^{-1})^\top$ we have
$\boldsymbol{\alpha\beta}^\top=\boldsymbol{\alpha}^*(\boldsymbol{\beta}^*)^\top$.
If cointegration exists, then equation (\ref{eq:vec}) can be written
as follows:
\begin{equation}\label{eq:vecfull}
\Delta\mathbf{y}_t 
= \boldsymbol{\alpha\beta}^\top\mathbf{y}_{t-1} 
  + \sum_{i=1}^{p-1}\boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
\end{equation}
\noindent
which is a VAR model but for time series differences.

Transposing each equation of the system (\ref{eq:vecfull}) we can write
the VECM($p$) model in block-matrix form as:
\begin{equation}\label{eq:vareq}
\mathbf{B} = 
\mathbf{A} \mathbf{X} + 
\mathbf{E} \, , 
\end{equation}
%
\noindent where $\mathbf{B}$ dimension is $((N-p)\times l)$, $\mathbf{A}$
dimension is $((N-p)\times(r+(p-1)l +1))$, $\mathbf{X}$ dimension is $((r+(p-1)l
+1)\times l)$ and $\mathbf{E}$ dimension is $((N-p)\times l)$:
%
\begin{alignat}{3}
\mathbf{B}
&= \begin{bmatrix}
   \Delta\mathbf{y}_{p+1}^\top \\
   \Delta\mathbf{y}_{p+2}^\top \\
   \vdots \\
   \Delta\mathbf{y}_N^\top
   \end{bmatrix}
&\quad
\mathbf{X}
&= \begin{bmatrix}
   \boldsymbol{\alpha}^\top \\
   \boldsymbol{\Phi}_1^{*\top} \\
   \boldsymbol{\Phi}_2^{*\top} \\
   \vdots \\
   \boldsymbol{\Phi}_{p-1}^{*\top} \\
   \mathbf{c}^\top
   \end{bmatrix}
&\quad
\mathbf{E}
&= \begin{bmatrix}
   \boldsymbol{\epsilon}_{p+1}^\top \\
   \boldsymbol{\epsilon}_{p+2}^\top \\
   \vdots \\
   \boldsymbol{\epsilon}_N^\top \\
   \end{bmatrix}
\end{alignat}
\noindent and 
\begin{align}
\mathbf{A} 
&= \begin{pmat}[{....|}]
   \mathbf{y}_p^\top \boldsymbol{\beta} & \Delta \mathbf{y}_p^\top & \Delta\mathbf{y}_{p-1}^\top & \dots 
                    & \Delta\mathbf{y}_2^\top & 1 \cr
   \mathbf{y}_{p+1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{p+1}^\top & \Delta\mathbf{y}_p^\top & \dots
                       & \Delta\mathbf{y}_3^\top & 1 \cr
   \vdots & \vdots & \vdots & \ddots & \vdots & \vdots \cr
   \mathbf{y}_{N-1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{N-1}^\top & \Delta\mathbf{y}_{N-2}^\top & \dots 
                       & \Delta\mathbf{y}_{N-p-1}^\top & 1 \cr
   \end{pmat}\, .
\label{eq:Amatrix}
\end{align}
Taking into account the error term $\mathbf{E}$, equation~(\ref{eq:vareq}) 
can be solved with respect to $\mathbf{X}$ using the OLS estimation.

@@

Cointegration vectors can be found applying the Johansen method which uses a
sample of the last historical data. However, VECM assumes cointegration vectors
do not change in time.  In fact, \cite{gregoryETal1996} addresses that the
long-run relationships between the time series  might change due to several
economic factors that can lead to structural breaks in the cointegration
relationship.  In order to show that the number of cointegration vectors depends
on the amount $L$ of historical data and the number of lags $p$ in the VECM, we
used a grid search.  We arbitrarily defined a grid of possible values for $L$
and $p$. $L$ goes throughout $[2,14]$ hours (1 $hour = 360$ data points) with a
step size of 4 hours and $p$ throughout $[1,5]$ with a step size of 1. The idea
was to show the variability of the number of cointegration vectors when we changed
these two parameters. We used four forex rates: EURUSD, GBPUSD, USDCHF and
USDJPY with 10-second frequency. Data started at 13:00 GMT of the 13th of August
2014, when the New York and London financial markets opened.

\begin{figure}[!h]
  %\vspace{-0.8cm}
  \centering
   \includegraphics[width=0.9\textwidth]{img/51_Fig1}
  \caption{Distribution of the number of cointegration vectors using $p=1$ lags.
  Four possible values for windows size $L$ are shown (2, 6, 10 and 14) hours (1
  hour = 360 data points).}
  \label{fig:hists}
\end{figure}

Figure \ref{fig:hists} shows the distribution of the number of cointegration
vectors given by the Johansen method for different values of $L=[2,6,10,14]$
hours and $p=1$ . This procedure was carried out by a sliding window of
historical data moving 1000 times. We observed that the distribution of
cointegration vectors changed with different values of $L$. When
$L=2$ hours, there was no cointegration in more than 60\% of the iterations.
Cointegration increased when $L=6$ hours and was maximum when $L=10$ hours where
one cointegration vectors was found for all 1000 iterations. The ocurrence of
cointegration started to decrease at $L=14$ hours.

@@

The goal of our next experiment was to find a relationship between the ratio $PC$
and the performance measure MSE (see equation \ref{eq:MSE}). $L$ was defined
between $[2,14]$ hours, that corresponded to $[720, 5040]$ data points, and $p$
took values between $[1,5]$.

\begin{figure}[ht!]
  \centering
   \includegraphics[width=0.9\textwidth]{img/51_Fig2}
  \caption{MSE versus the percentage of cointegration considering 1000
  iterations. Optimum windows size $L$ found was 1060. Below MSE versus $L$ shows a
  rapidly decreasing behaviour, founding minimum at PC = 73\% }
  \label{fig:cointvsmse}
\end{figure}

Figure~\ref{fig:cointvsmse} shows the relationship between MSE and $PC$ and $L$. We
found that higher cointegration percentage leads to improved performance accuracy
in terms of lower MSE. Also, increasing the size of the sliding window $L$
doesn't necessarily help to reduce MSE.

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.9\textwidth]{img/51_Fig3}
  \caption{Computing time of sequential and parallel algorithm is shown in the
  upper figure. Speed-up is shown below.}
  \label{fig:extimes}
\end{figure}
